---
title: "Time_series analysis and forecasting"
author: Bahman Rostami-Tabar
execute:
  echo: false
  message: false
  warning: false
  cache: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  pdf:
    number-sections: true
    keep-tex: true
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: setup
#| include: false
library(tidyverse)
library(fpp3)
library(lubridate)
library(tsbox)
library(dygraphs)
```


# Prepare data for analysis and forecasting

## Read data

```{r}
#| label: arrival
ae_arrivals <- read_csv("data/AE_arrivals_hourly.csv")
```


## Check duplications and fix it

```{r}
#| label: duplicate
#2.1- 
n_duplicated <- ae_arrivals |> duplicated() |> sum()#check duplicates

# remove duplicates using dplyr::distinct()  and get a distinct tibble
```

## Create tsibble

```{r}
#| label: tsibble
arrival_tsb <- ae_arrivals |> 
  as_tsibble(index = hour, key= c(gender,type_injury))
arrival_tsb

```

You can also check for temporal gaps and if there is any, fix them.


## Create daily hospital arrivals with keys

```{r}
#| label: ae_daily_key
ae_arrival_daily <- arrival_tsb |> group_by(gender, type_injury) |> 
  index_by(date=lubridate::as_date(hour)) |> 
  summarise(arrival=sum(arrival), .groups = "drop")
ae_arrival_daily
```


## Create total hourly arrivals

```{r}
#| label: hourly
arrival_hourly <- arrival_tsb |> 
  index_by(hour) |> 
  summarise(arrival=sum(arrival), .groups = "drop")
arrival_hourly
```


## Create total daily arrivals

```{r}
#| label: daily
arrival_daily <- arrival_tsb |> 
  index_by(date=lubridate::as_date(hour)) |> 
  summarise(arrival=sum(arrival), .groups = "drop")
arrival_daily
```


```{r}
arrival_tsb |> group_by(gender, type_injury) |> 
  index_by(week=yearweek(hour)) |> 
  summarise(arrival=sum(arrival), .groups = "drop")

arrival_tsb |>
  index_by(week=yearweek(hour)) |> 
  summarise(arrival=sum(arrival), .groups = "drop")
```


```{r}
arrival_tsb |> group_by(gender, type_injury) |> 
  index_by(quarter=yearquarter(hour)) |> 
  summarise(arrival=sum(arrival), .groups = "drop")

arrival_tsb |>
  index_by(month=yearmonth(hour)) |> 
  summarise(arrival=sum(arrival), .groups = "drop")
```


# Visualise

## time plot

You can create time plot using `autoplot()` function

```{r}
#| label: ts-plot-hourly
arrival_tsb
arrival_tsb |> autoplot(arrival) # create a time plot of hourly data
```

write what do you observe:

```{r}
#| label: ts-plot-daily
arrival_daily |> autoplot(arrival) # create a time plot of daily data
```

write what do you observe:

```{r}
#| label: hourly-arrival-dygraph
tsbox::ts_xts(arrival_tsb) |> 
  dygraph() |> 
  dyRangeSelector(dateWindow = c("2010-01-01", "2010-01-02"))
```


write what do you observe:


```{r}
#| label: daily-arrival-dygraph
tsbox::ts_xts(arrival_daily) |> 
  dygraph() |> 
  dyRangeSelector(dateWindow = c("2010-01-01", "2010-02-01"))
```

write what do you observe:

## Seasonal plots

use seasonal and subseries plots to check whether series contain seasonality. You can also try this with hourly series, you can change the "period= ..." argument to see different periods of seasonality.

```{r}
#| label: seasonal-plot-hourly
arrival_hourly |> feasts::gg_season(arrival, period = "week") #seasonal plot
```

```{r}
#| label: seasonal-plot-daily
arrival_daily |> feasts::gg_season(arrival, period = "week") #seasonal plot
```

## subseries plots

```{r}
#| label: sub-series
arrival_hourly |> feasts::gg_subseries(arrival, period = "day")# subseries plot
arrival_daily |> feasts::gg_subseries(arrival,period = "week")# subseries plot
```

Is there any seasonality in the hourly and daily time series?

## autocorrelation plot

```{r}
#| label: acf-calcul
arrival_daily |>
  feasts::gg_lag(arrival, lags = 1:21, geom = "point")# create lag plots for 14 lags, from 1 to 14

arrival_daily |> ACF(lag_max = 21) |> autoplot()# plot acf# compute autocorrelation function for 14 lags and plot then
```

# Specify and train models
## Time series cross validation (rolling origins)

This is also called rolling forecast or rolling origin:
You can also reflect on the following questions:
- Why do we use TSCV? you can read more here: https://otexts.com/fpp3/tscv.html
- How do we do TSCV in R? Which steps to follow?
      1. split data using `filter_index()`
      2. create different time series (different origins)
      3. model each time series, 
      4. forecast for each series
      5. evaluate the forecast accuracy

let's see how we do it in R:

## split data

We initially split the data into test and train, the size of test set equals the forecast horizon, we use this for the purpose of visualisating the forecasts, not deciding which model is the best(more accurate)

```{r}
#| label: split
#| include: false
f_horizon <- 42# forecast horizon
percentage_test <- 0.2 #20% of time series for test set

test <- arrival_daily |> 
  filter_index(as.character(max(arrival_daily$date)-round(percentage_test*length(unique(arrival_daily$date)))+1) ~ .)

train <- arrival_daily |>
  filter_index(. ~ as.character(max(arrival_daily$date)-(round(percentage_test*length(unique(arrival_daily$date))))))
```

## Use `stretch_tsibble()` to Create different timeseries (i.e different id)

We apply time series cross validation on the train data. We first start with an initial training size (.init = ) and then increase the size of the previous time series by adding more new observation(.step=) to create a new time series, we continue creating these timeseries until the number of observation left at the end of timeseries equals to the forecast horizon, we stop there.

```{r}
#| label: rolling-origin-series
train_tscv <- arrival_daily |> 
  filter_index(. ~ as.character(max(arrival_daily$date)-(f_horizon))) |>
  stretch_tsibble(.init = length(unique(train$date)), .step = 7) # split data into different time series (i.e. origin or id) with increasing size
train_tscv
```

## train models for each time series

```{r}
#| label: train-model
ae_model_tscv <- train_tscv |>
model(
    mean = MEAN(arrival),#total average
    naive = NAIVE(arrival),#naive
    snaive = SNAIVE(arrival),#seasonal naive
    exponential_smoothing = ETS(arrival),#exponential smoothing naive
    arima = ARIMA(arrival),#ARIMA,
    ar = AR(arrival),#AR
    regression = TSLM(arrival ~ trend() + season()),#regression
  )
ae_model_tscv
```

# Generate forecast

## Forecast for each series

Now, we want to produce forecast for 42 days that is applied to all time series created using TSCV:

```{r}
#| label: fcst-series
ae_fcst_tscv <-  ae_model_tscv |> 
  forecast(h=f_horizon)
ae_fcst_tscv #observe ae_fcst_tscv
```

## Extract prediction intervals

Use `hilo()` to extract prediction intervals for any coverage probability you are interested in. To be able to see values for lower bound and upper bound in separate columns, you need to unpack the prediction intervals using `unpack_hilo()`:

```{r}
#| label: prediction-interval-extract
## getting prediction intervals
ae_fcst_tscv |> hilo(level = 95) |> unpack_hilo(`95%`)
```

# Evaluate forecast accuracy

You calculate the point forecast accuracy using `accuracy()` function. `accuracy()` needs both the forecast object(fable) and actual data.
This will provide a summary of multiple accuracy measures. The result is summarised automatically across all series (.id) using a simple average.

```{r}
#| label: all-accuracy-measures
#point_accuracy_measures, interval_accuracy_measures, o and distribution_accuracy_measures)
fc_accuracy <- ae_fcst_tscv |> 
  accuracy(arrival_daily,
           measures = list(point_accuracy_measures,
                           interval_accuracy_measures,
                           distribution_accuracy_measures
))

fc_accuracy |> select(.model,MASE,RMSSE,winkler,CRPS)

```


# Forecast for the future using the best model

Once you compare the forecast accuracy and find out which model produces the most accurate forecadt, you can apply it to thw whole data to forecast the future:

```{r}
#| label: forecast-future
forecast_future <- arrival_daily |> model(regression = TSLM(arrival ~trend()+season())) |>  forecast(h=f_horizon)

forecast_future |> autoplot(arrival_daily |> filter_index("2016" ~ .))
```

## Produce probabilistic forecast using bootstrapping

Most time series models produce normally distributed forecasts — that is, we assume that the distribution of possible future values follows a normal distribution. 

When a normal distribution for the residuals is an unreasonable assumption, one alternative is to use bootstrapping, which only assumes that the residuals are uncorrelated with constant variance. 

You can also use `generate()` function to generate futures using bootstrapping:


```{r}
#| label: bootstrap-generate

fit <- arrival_daily |>
  model(autoregressive=ETS(arrival))

sim <- fit |> generate(h = 42, times = 5, bootstrap = TRUE)
sim
```


You can use `forecast()` directly to generate futures:

```{r}
#| label: bootstrap-forecast
arrival_fc <- fit |>
  forecast(h = f_horizon, bootstrap = TRUE, times = 5)
arrival_fc
```





